Analysis and Figures for Gallego et al, 2020

### 2020-09-26 Manuscript/Response.to.reviewers.tex

Figure 2 should be exactly the same as the first part of the Supplemental figure - A thorough review of the code that generates the supplemental figures revealed a small difference in the input dataset. We thank the reviewer for noticing this, and now the code that generates each of the figures is consistent.

\item{As Figure 3 is divided into sections A, B, and C, these ought to be referred to in the text (see below). The text is very small and hard to read in the review version I have.}

\textbf{Response:}

In an effort to declutter Figure 3, and following suggestions from reviewer 1, we have focused the figure in those species showing a more dramatic change in suitability. The new version of the figure has three panels (A, B, C; one per phylum) displaying three aspects of the data (.1, .2 and .3) and we refer to them specifically in the text

@@

\item{Line 99: The value of predicting which cluster (?community?) a given set of environmental parameters will fall within is not entirely clear to me. Is there a consistent set of taxa that make of up these clusters across time and space? Can specific groups of taxa be deduced from the ?community? number? Do the environmental variables accurately predict which set of taxa will be present? This section needs a bit more clarity.}

\textbf{Response:}

In sum, yes: these clusters are coherent groups of species and predictably occur in different environmental conditions.  We have revised the main text to be more clear in this regard.

{\color{red}RYAN: How about this:
In sum, yes: these clusters are coherent groups of species and predictably occur in different environmental conditions. However, differences in relative proportions of species will also contribute to cluster dissimilarities. }

@@

We think the relationship between clusters and the environmental variables is important in a few aspects. First, as a means of ground truthing our findings, it has great value as the clusters were detected blind to the environmental data, and the CAP analysis return the distinctiveness of these clusters in that space - Figure 2B highlights the translation of those modified axis of the CAP plot into the real values of temperature and pH. Second, these clusters (maybe communities is a more committed term that requires a longer dataset in which to confirm co-ocurrence of members) do indeed reflect changes in the relative importance of the species present.

\item{Line 102-104: According to my interpretation of Fig. 2C, it seems like the colder water communities actually had *lower* values of the given taxa, as the differences are all negative relative to other communities. It may help to specify in the text which characteristics are specific to which communities (communities 2 and 3 both span the lower temperature area of Fig. 2B).}

 Figure 2C includes now taxa with the higher contribution to the dissimilarity between clusters, highlighting both positive and negative contributions.

@@

\textbf{Response:}

Done.

\item{Line 121-133: When referring to specific taxa as winners or losers, I was expecting to see them in Figure 3B, but not all of them were there. Or if it did appear, it was only at one site, but it was mentioned as being an overall winner or loser.}

 We thank the reviewer for clarifying this point. We have now decluttered Figure 3; and highlight taxa that are now mentioned in the text: the Dinoflagellates Karlodinium, Gymnodinaceae and Alexandrium; the Diatoms Stephanodiscus, Nitzschia, Coscinodiscus and Chaetoceros; and the Haptophytes Emiliania and Chrysochromulina spp.

@@

 We thank the reviewer for clarifying this point. We have now decluttered Figure 3; and highlight taxa that are now mentioned in the text: the Dinoflagellates Karlodinium, Gymnodinaceae and Alexandrium; the Diatoms Stephanodiscus, Nitzschia, Coscinodiscus and Chaetoceros; and the Haptophytes Emiliania and Chrysochromulina spp.

\item{Line 122-123: According to Figure 3B, Thalassionema, Navicula, and Nitzschia appear to refer specifically to San Juan. If so, please state this.}

Our effort has been on visualizing species with more pronounced change, not to highlight species that respond in contrasting ways in the San Juan Islands and the Hood Canal. Most taxa react in the same direction in both regions; with a few exceptions in which species are almost absent on one study area.

@@

Our effort has been on visualizing species with more pronounced change, not to highlight species that respond in contrasting ways in the San Juan Islands and the Hood Canal. Most taxa react in the same direction in both regions; with a few exceptions in which species are almost absent on one study area.

\item{Line 123: Coscinodiscus and Ditylum seem to refer to Hood Canal. This ought to specified as before.}

Coscinodiscus sees its probability of presence reduced on both Areas. The same happens for D. brightwelli, although our approach for a more readable figure has seen this taxa drop into the Supplementary figures

\item{Line 129: Write out Emiliania. The 'sp.' after Alexandrium should not be italicized. This sentence might read better if written: 'The coccolithophore Emilinania huxleyi and the dinoflagellate Alexandrium sp. both find more suitable habitat' (rather than 'each finds')}

 We have updated that sentence, which now reads: "The haptophytes \textit{Emiliania huxleyi} and \textit{Chrysochromulina} sp. the dinoflagellate \textit{Alexandrium} sp. all find more suitable habitat in both of our study areas." Thanks for the suggestion.

@@

\textbf{Response:}

 We have updated that sentence, which now reads: "The haptophytes \textit{Emiliania huxleyi} and \textit{Chrysochromulina} sp. the dinoflagellate \textit{Alexandrium} sp. all find more suitable habitat in both of our study areas." Thanks for the suggestion.

\item{Line 136-137: This is minor, but this section (or methods) might benefit from a brief mention of why ANOVA was used in one case and Wilcoxon in another.}

We thank the reviewer for pointing this out: We have rethought this and applied Wilcoxon tests for both global and phylum-specific changes in diversity. As we are comparing pairs of data we prefer not to assume normal distributions of the datasets or the differences between them, Wilcoxon test was a better choice.

@@

We thank the reviewer for pointing this out: We have rethought this and applied Wilcoxon tests for both global and phylum-specific changes in diversity. As we are comparing pairs of data we prefer not to assume normal distributions of the datasets or the differences between them, Wilcoxon test was a better choice.

\item{Line 138-141: It seems like this paragraph could be moved to the discussion section.}

 A reworded version of this paragraph is now at the end of the discussion section.

@@

 A reworded version of this paragraph is now at the end of the discussion section.

\item{Line 147-148: Need a citation for Coscinodiscus as key fish food.}

 Agreed: We now cite the work of J. Zamon (2002) and S. Pasquaud and colleagues (2010) to support the role of Coscinodiscus in the food chain.

@@

\textbf{Response:}

  Done.

\item{Line 192: Might be helpful to specifically mention carbon sequestration as an important aspect of sinking rates.}

 Agreed - it will definitely have an effect on the `fate` of the photosynthesized carbon- a quickly sunk carbon will join marine sediments instead of being recycled in the upper layers of the water masses. We now cite the work of Polimene and colleagues on the role of phytoplankton and carbon sequestration.

@@

 Agreed - it will definitely have an effect on the `fate` of the photosynthesized carbon- a quickly sunk carbon will join marine sediments instead of being recycled in the upper layers of the water masses. We now cite the work of Polimene and colleagues on the role of phytoplankton and carbon sequestration.

\item{Line 204-205: Citation needed for fish mortality. Eutrophication seems like a cause of blooms, not a result of them. What are the potential cascade effects on benthic-pelagic coupling? The term here on its own isn?t really saying anything.}

 In their work, D. Anderson and colleagues cite a range of consequences of increased HAB blooms, including fish mortality and anoxia - and not eutrophication as we wrongly pointed out. Thanks for noticing. On the second argument - we saw potential consequences in two ways: by affecting recruiting of benthic species with pelagic larvae with reduced diatom diversity/biomass; and by changing the nature of sinking matter for filter feeders. We realize that bentho pelagic coupling is not the only source of carbon for benthic communities in the area of study so we have decided to remove this assertion from the discussion.

\item{Line 208-214: This is a big aspect of why this study is valuable! I think it's worth discussing this in the introduction.}

 We thank the reviewer for this suggestion.

@@

\item{Line 241: Should be Figure 1C}

{\color{red} Done }

\item{Line 267-268: Which index of abundance from Kelly et al. (2019) was used? That paper presents several and though it declares one to be more useful, I think it would be helpful in this manuscript to state which one was used.}

We used the eDNAindex (the proportion of reads assigned to each taxa per sample, scaled to the maximum proportion that taxa achieved). We provide the R implementation of this index in the Supplementary material, and have noted so in the text (line 128).

\item{Line 273-282: It's hard to discern based on the limited code and data provided exactly how the authors subsampled the different subregions of Puget Sound, since the paper they base their data on has overall values for the whole area. I think this section could use a bit of clarification, and all the code and raw data should be given (the provided github link is broken).}

We apologize for the broken github code - We have clarified how we simulated the future projection in the Supplementary Information - Also see response to reviewer1 on line 122 - 137 of this letter.

### 2020-09-26 Manuscript/What.lies.ahead.reviewed.tex
YES IT IS DONE

\textbf{Ram\'{o}n Gallego \textsuperscript{$a,b,1$}, Emily Jacobs-Palmer \textsuperscript{$b$}, Kelly Cribari \textsuperscript{$b$}, Ryan P. Kelly}\textsuperscript{$b$}
\vspace{1em}

\textbf{$^a$} {\small NRC Research Associateship Program,
Northwest Fisheries Science Center,
National Marine Fisheries Service,
National Oceanic and Atmospheric Administration. 2725 Montlake Boulevard East, Seattle, WA 98112}

\textbf{$^b$} {\small School of Marine and Environmental Affairs. University of Washington. 3707 Brooklyn Ave NE Seattle, WA 98105}

@@

We collected water samples to assess eDNA communities in two regions of the Salish Sea (Washington, USA): San Juan Island and the Hood Canal. These sites experience substantial variation in water chemistry and other environmental conditions despite geographic proximity (ca. 300km; Fig. 1). We sampled eight sites monthly for approximately 1.5 years (March 2017 to August 2018), taking three 1L samples (biological replicates; ca. 10m apart) each month at each site (261 bottle samples total). Each sample was filtered through a 0.45 $\mu$m  cellulose filter, and the filter preserved in Longmire buffer until DNA extraction \cite{renshaw2015room}. Concurrently, we collected one 120 ml water sample from each site and poisoned it with 0.1 ml of saturated HgCl$_2$ for carbonate chemistry analysis, following \cite{riebesell2011guide}. We also collected \textit{in situ} measurements of temperature, salinity and dissolved oxygen using a handheld multiprobe (Hanna Instruments, USA) and a portable refractometer. We note that many unmeasured variables influence planktonic communities (e.g., nutrients, sunlight, wave energy), but that our set of measured parameters clearly distinguished communities and was adequate for our purposes.

@@

Metabarcoding analysis of eDNA samples generated more than 50.8M sequences from 778 samples. These samples represented biological and technical replicates from 86 unique sampling events. After bioinformatics quality-control the dataset included $\sim${}45M sequences, from 4849 unique amplicon sequence variants (ASVs). Of these, 1364 ASVs (22.6M reads) could be annotated to a taxonomic level of Family or lower. These $\sim$ 500 taxa from 43 phyla were split according to their natural history and habitat (benthic vs. planktonic; see Supplementary Information). Because we expect planktonic taxa to vary with water mass \cite{kelly2018tides} and therefore with bottle-sampled carbonate chemistry, here we focus on only the planktonic taxa (N = 221). These taxa showed a seasonal richness gradient between study areas, consistent with documented biodiversity clines in the area (Fig. 1B.1) \cite{dethier_decadal_2008}.

Bray-Curtis dissimilarities among samples revealed large differences in metabarcoding communities due to geographic Area (Hood Canal vs. San Juan Island; F = 1.6184; p $<$ 0.01). We therefore performed a constrained analysis of principal components (CAP) for samples within each Area, showing the differences among communities as a function of temperature, pH, and salinity (Fig. 2; for clarity, results for Hood Canal shown; full analysis in the Supplementary Information).

 Each biological cluster (colored hulls, Fig. 2A) occupied a unique area of environmental parameter space. Planktonic communities therefore varied predictably with water temperature, salinity, and pH, across a range of those parameters likely to be encountered in many near-term future-ocean scenarios \cite{murray2015inland}. 
 Multinomial logistic regression yielded predictions of the most-likely community for any combination of environmental parameters (Fig. 2B).

These communities were distinguished by changes in the relative abundances of a wide variety of taxa. In the Hood Canal for example, the community linked with colder water and higher-pH (community 2 in Fig. 2C) showed higher eDNA indexes of diatoms (\textit{Minutocellus polymorphus}), green algae  (\textit{Bathycoccus prasinos} and \textit{Micromonas pusilla}), and dinoflagellates like \textit{Karlodinium} sp relative to the lower-pH cluster at the same temperature. In that community (cluster 1) \textit{Emiliania huxleyi}, \textit{Ditylum brightwellii} and the copepod \textit{Paracalanus} sp. C are more prevalent.  The community common in the high temperature range (cluster 3 in Fig. 2C) shows high values of \textit{Phaeocystis globosa},\textit{Ditylum brightwellii} among other species. Cluster 3 (Fig. 2A,B) and similar planktonic communities occupy the spectrum of environmental conditions most likely to be encountered in near-future climate scenarios as temperature rises and pH falls. For example, we expect the Hood Canal in 2095 to have the conditions in which community 3 is the most likely community 66\% of the time, an increase of 11\% compared to 2017(see Supplementary Information). On the other hand, conditions that make cluster 2 the most likely community drop from 9\% to 1\% of the time between 2017 and 2095.

@@

To explore the suitability of different environmental conditions for each taxon, we modeled the likelihood of taxon presence as a function of temperature, pH, and geographic Area as described in the Methods. Salinity was not informative in our models, as it is highly correlated with temperature in our dataset and is moreover predicted to remain largely unchanged in future scenarios \cite{khangaonkar2019salish}. Model projections let us show the change in the probability of presence of each individual taxon for 2095 vs 2017 (Fig. 3 A.1, B.1, C.1), estimate richness for larger taxonomic grounds as a whole for these two climates (Fig. 3 A.2, B.2, C.2), and estimate richness within taxonomic groups across the pH-temperature continuum (Fig. 3 A.3, B.3, C.3).

@@

Likely winners under future conditions are more widely scattered among higher taxonomic groups. The haptophytes \textit{Emiliania huxleyi} and \textit{Chrysochromulina} sp. the dinoflagellate \textit{Alexandrium} sp. all find more suitable habitat in both of our study areas. Among others not shown in Fig 3, \textit{Chaetoceros} (diatom) and many hydrozoans (Cnidaria) likely increase in San Juan Island, and the potentially fish-killing heterokont flagellate \textit{Pseudochattonella} increases in both study areas. See Supplementary Information for a complete list of taxon-specific projections.

@@

Temperate surface oceans worldwide average approximately 14$\degree$C and pH of 8.1 \cite{bopp2013multiple}, and will change substantially in this century \citep[mean $\Delta$ T 2.5$\degree$C, $\Delta$ pH -0.35 globally; RCP 8.5;][]{gattuso2015contrasting}.  Here we document communities exposed to this same range of projected conditions in the present day, along an environmental gradient only ca. 200km wide, allowing us to project future ocean communities from a robust set of underlying observations. Our results reflect patterns in a diverse selection of species from nearshore marine communities in the Salish Sea, consisting of 222 planktonic taxa obtained from the metabarcoding analysis of 227 discrete samples across 77 space-time points (eight sites, 1.5 years). We find that changes in the composition of biological communities closely mirrored the variation in pH and temperature, with clear winners (e.g., \textit{Emiliania huxleyi}, \textit{Alexandrium}, and others) and losers (many, but not all, diatoms) likely to shift the structure and function of future marine communities.

@@

 We are grateful to the UW Center for Environmental Genomics and the Conservation Biology Molecular Genetics Laboratory at NOAA Fisheries (Montlake) for access to lab equipment and associated expertise and support. Thanks to Micah Horwith and the Washington State Department of Natural Resources for in-field support, to Terrie Klinger for feedback on an earlier draft, to Stephanie Moore for advice on phytoplankton generally, to Ana Ram\'{o}n-Laca for lab help, to Dan Drinan for field help, and to Alex Gagnon for access to carbonate-chemistry expertise and equipment. The comments of two reviewers helped us improve the manuscripts and encourage us to streamline the analysis of our data.

>>>>

\textbf{Ram\'{o}n Gallego \textsuperscript{$a,b,1$}, Emily Jacobs-Palmer \textsuperscript{$a$}, Kelly Cribari \textsuperscript{$a$}, Ryan P. Kelly}\textsuperscript{$a$}
\vspace{1em}

\textbf{$^a$} {\small School of Marine and Environmental Affairs. University of Washington. 3707 Brooklyn Ave NE Seattle, WA 98105}

\textbf{$^b$} {\small Current affiliation: NRC Research Associateship Program,
Northwest Fisheries Science Center,
National Marine Fisheries Service,
National Oceanic and Atmospheric Administration. 2725 Montlake Boulevard East, Seattle, WA 98112}

@@

We collected water samples to assess eDNA communities in two regions of the Salish Sea (Washington, USA): San Juan Island and the Hood Canal. These sites experience substantial variation in water chemistry and other environmental conditions despite geographic proximity (ca. 300km; Figure 1). We sampled eight sites monthly for approximately 1.5 years (March 2017 to August 2018), taking three 1L samples (biological replicates; ca. 10m apart) each month at each site (261 bottle samples total). Each sample was filtered through a 0.45 $\mu$m  cellulose filter, and the filter preserved in Longmire buffer until DNA extraction \cite{renshaw2015room}. Concurrently, we collected one 120 ml water sample from each site and poisoned it with 0.1 ml of saturated HgCl$_2$ for carbonate chemistry analysis, following \cite{riebesell2011guide}. We also collected \textit{in situ} measurements of temperature, salinity and dissolved oxygen using a handheld multiprobe (Hanna Instruments, USA) and a portable refractometer. We note that many unmeasured variables influence planktonic communities (e.g., nutrients, sunlight, wave energy), but that our set of measured parameters clearly distinguished communities and was adequate for our purposes.

@@

Metabarcoding analysis of eDNA samples generated more than 50.8M sequences from 778 samples. These samples represented biological and technical replicates from 86 unique sampling events. After bioinformatics quality-control the dataset included $\sim${}45M sequences, from 4849 unique amplicon sequence variants (ASVs). Of these, 1364 ASVs (22.6M reads) could be annotated to a taxonomic level of Family or lower. These $\sim$ 500 taxa from 43 phyla were split according to their natural history and habitat (benthic vs. planktonic; see Supplementary Information). Because we expect planktonic taxa to vary with water mass \cite{kelly2018tides} and therefore with bottle-sampled carbonate chemistry, here we focus on only the planktonic taxa (N = 221). These taxa showed a seasonal richness gradient between study areas, consistent with documented biodiversity clines in the area (Figure 1B.1) \cite{dethier_decadal_2008}.

Bray-Curtis dissimilarities among samples revealed large differences in metabarcoding communities due to geographic Area (Hood Canal vs. San Juan Island; F = 1.6184; p $<$ 0.01). We therefore performed a constrained analysis of principal components (CAP) for samples within each Area, showing the differences among communities as a function of temperature, pH, and salinity (Figure 2; for clarity, results for Hood Canal shown; full analysis in the Supplementary Information).

 Each biological cluster (colored hulls, Figure 2A) occupied a unique area of environmental parameter space. Planktonic communities therefore varied predictably with water temperature, salinity, and pH, across a range of those parameters likely to be encountered in many near-term future-ocean scenarios \cite{murray2015inland}. 
 Multinomial logistic regression yielded predictions of the most-likely community for any combination of environmental parameters (Figure 2B).

These clusters were distinguished by changes in the relative abundances of a wide variety of taxa. In the Hood Canal for example, the cluster linked with colder water and higher-pH (cluster 2 in Figure 2B) showed higher eDNA indexes of diatoms (\textit{Minutocellus polymorphus}), green algae  (\textit{Bathycoccus prasinos} and \textit{Micromonas pusilla}), and dinoflagellates like \textit{Karlodinium} sp. relative to the lower-pH cluster at the same temperature range. In that community (cluster 3) \textit{Emiliania huxleyi}, \textit{Ditylum brightwellii} and the copepod \textit{Paracalanus} sp. C are more prevalent.  The cluster common in the high temperature range (cluster 1 in Figure 2B) shows high values of \textit{Phaeocystis globosa},\textit{Emiliania huxleyi}, \textit{Ditylum brightwellii} among other species. Cluster 1 (Figure 2A,B) and similar planktonic communities occupy the spectrum of environmental conditions most likely to be encountered in near-future climate scenarios as temperature rises and pH falls. For example, we expect the Hood Canal in 2095 to have the conditions in which community 1 is the most likely community 67\% of the time, an increase of 11\% compared to 2017(see Supplementary Information). On the other hand, conditions that make cluster 2 the most likely community drop from 12\% to 2\% of the time between 2017 and 2095.

@@

To explore the suitability of different environmental conditions for each taxon, we modeled the likelihood of taxon presence as a function of temperature, pH, and geographic Area as described in the Methods. Salinity was not informative in our models, as it is highly correlated with temperature in our dataset and is moreover predicted to remain largely unchanged in future scenarios \cite{khangaonkar2019salish}. Model projections let us show the change in the probability of presence of each individual taxon for 2095 vs 2017 (Figure 3 A.1, B.1, C.1), estimate richness for larger taxonomic grounds as a whole for these two climates (Figure 3 A.2, B.2, C.2), and estimate richness within taxonomic groups across the pH-temperature continuum (Figure 3 A.3, B.3, C.3).

@@

Likely winners under future conditions are more widely scattered among higher taxonomic groups. The haptophytes \textit{Emiliania huxleyi} and \textit{Chrysochromulina} sp. the dinoflagellate \textit{Alexandrium} sp. all find more suitable habitat in both of our study areas. Among others not shown in Figure 3, \textit{Chaetoceros} (diatom) and many hydrozoans (Cnidaria) likely increase in San Juan Island, and the potentially fish-killing heterokont flagellate \textit{Pseudochattonella} increases in both study areas. See Supplementary Information for a complete list of taxon-specific projections.

@@

Temperate surface oceans worldwide average approximately 14$\degree$C and pH of 8.1 \cite{bopp2013multiple}, and will change substantially in this century \citep[mean $\Delta$T 2.5$\degree$C, $\Delta$pH -0.35 globally; RCP 8.5;][]{gattuso2015contrasting}.  Here we document communities exposed to this same range of projected conditions in the present day, along an environmental gradient only ca. 200km wide, allowing us to project future ocean communities from a robust set of underlying observations. Our results reflect patterns in a diverse selection of species from nearshore marine communities in the Salish Sea, consisting of 222 planktonic taxa obtained from the metabarcoding analysis of 227 discrete samples across 77 space-time points (eight sites, 1.5 years). We find that changes in the composition of biological communities closely mirrored the variation in pH and temperature, with clear winners (e.g., \textit{Emiliania huxleyi}, \textit{Alexandrium}, and others) and losers (many, but not all, diatoms) likely to shift the structure and function of future marine communities.

@@

 We are grateful to the UW Center for Environmental Genomics and the Conservation Biology Molecular Genetics Laboratory at NOAA Fisheries (Montlake) for access to lab equipment and associated expertise and support. Thanks to Micah Horwith and the Washington State Department of Natural Resources for in-field support, to Terrie Klinger for feedback on an earlier draft, to Stephanie Moore for advice on phytoplankton generally, to Ana Ram\'{o}n-Laca for laboratory help, to Dan Drinan for field help, and to Alex Gagnon for access to carbonate-chemistry expertise and equipment. The comments of two reviewers helped us improve the manuscripts and encourage us to streamline the analysis of our data.

### 2020-09-26 Scripts/Manuscript.Main.Analysis.Rmd
YES IT IS DONE

@@

 left_top %>% set_names(Clusters.analysis.and.cap$Area) %>% map(~ .x +  theme_light() + theme(legend.position = "none") ) -> left_top

@@

testquick %>% pull(plots) %>% set_names(testquick$Area) %>% map(~ .x +  theme(axis.title.y = element_text(angle = 0),legend.position = "none") )-> center_top
```

>>>>

    set.seed(1231)

@@

 left_top %>% set_names(Clusters.analysis.and.cap$Area) %>% map(~ .x +  theme_light() + theme(legend.position = "none") ) -> left_top
 left_top

@@

testquick %>% pull(plots) %>% set_names(testquick$Area) %>% map(~ .x +  theme(axis.title.y = element_text(angle = 0),legend.position = "none") )-> center_top

center_top
```
